---
title: "Practical Fall 2019"
author: "Kristi Gdanetz MacCready"
date: "10/28/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Documents/UMich_Bio201_F19/Practial")
```

As with each lab, add "completed" to the Rmd filename; update the name in the author line above; update the location of the working directory as necessary. 

# Load packages
```{r Load packages, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(readxl)
library(broom)
library(cowplot)
library(ggplot2)
library(ggpubr)
library(stringr)
set.seed(7)
```

# Part 1 - concepts 
Type the answers to questions 1-5 in the space below. 

### Question 1

Given vector v1 below, will it be read as character or numeric? Why does this happen? [1 point]
```{r}
v1 <- c("one", "2", "three", "4")
typeof(v1)
```

It is read as character because there is a hierarchy of data types within the system, with character coming on top. 

### Question 2

dataframe1 has 323 rows and dataframe2 has 345 rows. After an inner join the resulting data fame has 312 rows. Why does the result have fewer rows? [1 point]

The function inner_join() retains all rows where the values are matching and includes all columns, differing or same accordingly. This would mean that any row in which a value is different would be excluded on the combined table. 

### Question 3

What are the three assumptions for a t-test? Explain what to do if the variance assumption is violated. Explain what to do if the normality assumption is violated. [3 points]

The three assumptions for a t-test or parametric test are:
-Adequate sample size (preferably n > 30)
-(Relative) Normality around the mean
-Equal variances between groups being examined

If the normality assumption is violated, then the associated non-parametric or Wilxocon Summed Ranks test must be used to evaluate the statistical difference in means. If equal variances are violated, then a paired t-test could potentially  still be performed. However, the data may need to be transformed, possibly logarithmically, to account for this difference in variances. If the variance-test p-value is close enough to the significance cutoff, the t-test could still just be applied with no further work done to it. If all else fails even afterwards, then it would be best to move on to the Wilcoxon and assume non-parametric conditions (or Mann-whitney if samples are independent). In the end, the most important requirement is that of normality. 

### Question 4

You have a data frame (called butyrate_df) with the columns: "Participant_ID", "Study_week", "Semester", "Supplement_consumed", "Avg_butyrate". If this data frame is piped to each of the functions below, evaluate if the description of the results is true or false. [3 points]
```{r eval=FALSE, include=FALSE}
butyrate_df %>%
  filter(Semester == "Fall2019")
# Result1 = keeps only data from Fall 2019 semester

butyrate_df %>%
  filter(Avg_butyrate <= 47.0)
# Result2 = keeps only values equal to 47 mmol/kg 

butyrate_df %>%
  group_by(Semester) %>%
  summarize(New_column = median(Avg_butyrate))
# Result3 = new_column contains the median of each participant 
```

Result 1 would be true, as the filter() function only filters out the data indicated by the argument to show up when run. Result 2 would not be completely true as the argument indicates an "equal to or less than" sign, meaning that anything less than 47 mmol/kg would be included as well, not just the values equal to 47. Result 3 is also false as new_column would give the Avg_butyrate median for each semester, not for each person. To get each participant median, it would have to be done, prior to the summarize(), as "group_by(Participant_ID)".

### Question 5

something about NAs

The data frame (called breath_df) has the columns: "Participant_ID", "Study_week", "Semester", "Supplement_consumed", "Methane", "Hydrogen", "Carbon_dioxide". This data frame is used for the code below, explain what will happen to any NAs in each of the shown calculations, and how the NAs will effect the underlying data. [2 points]
```{r eval=FALSE, include=FALSE}
# Option 1
breath_df %>%
  filter(!is.na(Methane)) %>% 
  group_by(Participant_ID, Study_week) %>%
  summarize(Avg_methane = mean(Methane)) 

# Option 2
breath_df %>%
  group_by(Participant_ID, Study_week) %>%
  summarize(Avg_methane = mean(Methane, na.rm = TRUE))
```

For Option 1, the !is.na() calls forth anything that is not an NA value within the "Methane" column. The group_by() and summarizze() lines the call forth a new calculation that will calculate the mean methane concentration by participant ID by study week, so for example, there would be a calculation for hypothetical "Participant 1" for both "week1"" and "week3". The "NA" values would thus be filtered out. 

For Option 2, the same would happen as the "na.rm = TRUE" argument excludes any NAs from the calculation, thus calling forth the average methane concentrations from only is.numeric positive values. 

# Part 2 - tidy data 

### Question 6

Find the data frame matching your unique name in the GitHub repository. Find the breath gas and taxa abundance data frames. Download and import all into RStudio and use these data frames to complete the rest of the practical. Name the dataframe to match your uniqname, name the breath gas data frame "breath", name the taxa abundance data frame "taxa". [1 point]
```{r}
# data import code 
aaaayang <- read_delim("Practial/Raw_data/aaaayang.txt",
                            delim = "\t", escape_double = FALSE, 
                            trim_ws = TRUE, na=c("NA"),
                            col_types = list())

breath <- read_delim("Practial/Raw_data/breath.txt",
                            delim = "\t", escape_double = FALSE, 
                            trim_ws = TRUE, na=c("NA"),
                            col_types = list())

taxa <- read_delim("Practial/Raw_data/taxa.txt",
                            delim = "\t", escape_double = FALSE, 
                            trim_ws = TRUE, na=c("NA"),
                            col_types = list())
```

Update column names in all files match the course conventions. [1 point]
```{r}
aaaayang <- aaaayang %>%
  rename_all(tolower) 
View(aaaayang)

breath <- breath %>%
  rename_all(tolower) 
View(breath)

taxa <- taxa %>%
  rename_all(tolower) %>%
  rename(relative_abundance = fract)
View(taxa)
```

View the supplement types in your semester. Are they all entered in the same style? If not update them to match. [1 point]
```{r}
#All supplement types are entered in the same style within my aaaayang data frame, but I will execute the code below just in case anything isnt first-letter capitalized: 

aaaayang[["supplement consumed"]] <- recode(.x=aaaayang[["supplement_consumed"]], 
                                  "accessible"="Accessible")
aaaayang[["supplement_consumed"]] <- recode(.x=aaaayang[["supplement_consumed"]], 
                                  "himaize"="HiMaize")
aaaayang[["supplement consumed"]] <- recode(.x=aaaayang[["supplement_consumed"]], 
                                  "inulin"="Inulin")
```

### Question 7

What are the dimensions of each data frame? Write the results as a comment after the function. [1 point]
```{r}
dim(aaaayang) #382 x 10
dim(taxa) # 46881 x 6
dim(breath) #2668 x 5
```

Select a join that will drop as few data points as possible. Join the data frame containing the participant data with the breath gas data. Name this "joined_df". [1 point]
```{r}
joined_df <- full_join(aaaayang, breath)
joined_df
```

What are the dimensions of the joined data frame? Write the results as a comment after the function. [1 point]
```{r}
dim(joined_df) #3846 x 12
#I noticed I have an extra supplement_consumed column in my "aaaayang" file, so maybe ncol should really be 11. 
```

### Question 8

Calculate the total SCFAs. [1 point]
```{r}
joined_df <- joined_df %>%
  mutate(total_scfa = acetate_mmol_kg + butyrate_mmol_kg + propionate_mmol_kg)
```

Calculate the weekly mean of each SCFA of each participant. Calculate the weekly mean methane and hydrogen levels of each participant. [1 point]
```{r}
joined_df <- joined_df %>%
  group_by(participant_id, study_week) %>%
  mutate(acetate_mean = mean(acetate_mmol_kg)) %>%
  mutate(butyrate_mean = mean(butyrate_mmol_kg)) %>%
  mutate(propionate_mean = mean(propionate_mmol_kg)) %>%
  mutate(ch4_mean = mean(ch4)) %>%
  mutate(h2_mean = mean(h2))
```

What are the dimensions of the data frame with the avearges? Write the results as a comment after the function.
```{r}
dim(joined_df) #3846 x 17
#I also included the total_scfa column.
```

Join this data frame with the one called "taxa". Name the new data frame "all_data". What are the dimensions of this new data frame? [1 point]
```{r}
all_data <- full_join(joined_df, taxa)
all_data
dim(all_data) #133581 x 21
```

Save all data frames generated to your GitHub page. [1 point]
```{r}
write_delim(aaaayang, path="Practial/Curated_data/aaaayang.txt",
            delim = "\t", quote = FALSE, col_names = TRUE)

write_delim(breath, path="Practial/Curated_data/breath.txt",
            delim = "\t", quote = FALSE, col_names = TRUE)

write_delim(taxa, path="Practial/Curated_data/taxa.txt",
            delim = "\t", quote = FALSE, col_names = TRUE)

write_delim(joined_df, path="Practial/Curated_data/joined_df.txt",
            delim = "\t", quote = FALSE, col_names = TRUE)

write_delim(all_data, path="Practial/Curated_data/all_data.txt",
            delim = "\t", quote = FALSE, col_names = TRUE)
```

# Part 3 - statistics & graphics 

Update all axes labels and figure titles so they are informative and easy to read. Upload final plots to GitHub. When specified, conduct the appropriate tests, and write the statistic (p, F, R value) as a comment. 

### Question 9

If you have more than one type of supplement consumed in your semester, filter for one type to answer Question 9. Plot each SCFA with the study week (weeks 1 and 3 only) on the x-axis. Make each study week a different color. Hint: you may have to use facets, or create individual plots and combine them into a multi-panelled figure. [3 points]
```{r}
aaaayang1 <- aaaayang %>%
  filter(supplement_consumed == "Inulin")
aaaayang1
  
acetate_plot <- ggplot(aaaayang1, aes(x=study_week, y=acetate_mmol_kg, color=study_week)) + 
  geom_violin() +
  xlab("Study Week") + 
  ylab("Acetate Concentration (mmol/kg)") 
acetate_plot
ggsave(path = "Figures", filename = "acetate_plot.png")

butyrate_plot <- ggplot(aaaayang1, aes(x=study_week, y=butyrate_mmol_kg, color=study_week)) + 
  geom_violin() +
  xlab("Study Week") + 
  ylab("Butyrate Concentration (mmol/kg)") 
butyrate_plot
ggsave(path = "Figures", filename = "butyrate_plot.png")

propionate_plot <- ggplot(aaaayang1, aes(x=study_week, y=propionate_mmol_kg, color=study_week)) +
  geom_violin() +
  xlab("Study Week") + 
  ylab("Propionate Concentration (mmol/kg)") 
propionate_plot
ggsave(path = "Figures", filename = "propionate_plot.png")

combined_plot <- ggarrange(acetate_plot, butyrate_plot, propionate_plot, 
          labels = c("Acetate", "Butyrate", "Propionate"),
          ncol = 3, nrow = 1)
combined_plot
ggsave(path = "Figures", filename = "combined_plot.png")

```

Based on the data presented in the plot you generated above, identify which statistical test you would need to determine if there was a change in SCFA concentration between weeks 1 and 3 of the study. Check the assumptions appropriate for the test. [3 points]
```{r}
#create week1 and week3 data frames: 
aaaayang_week1 <- aaaayang %>%
  filter(study_week == "week1") %>%
  filter(supplement_consumed == "Inulin") %>%
  mutate(total_scfa = acetate_mmol_kg + butyrate_mmol_kg + propionate_mmol_kg)
aaaayang_week1

aaaayang_week3 <- aaaayang %>%
  filter(study_week == "week3") %>%
  filter(supplement_consumed == "Inulin") %>%
  mutate(total_scfa = acetate_mmol_kg + butyrate_mmol_kg + propionate_mmol_kg)
aaaayang_week3

# assumptions: 
#Sample Size: 
aaaayang %>%
  filter(study_week == "week1") %>%
  filter(supplement_consumed == "Inulin") %>%
  summarize(sample_size = n()) # n = 82. n is greater than 30, so sample size is adequate.

aaaayang %>%
  filter(study_week == "week3") %>%
  filter(supplement_consumed == "Inulin") %>%
  summarize(sample_size = n()) # n = 74. n is greater than 30, so sample size is adequate.

#Normality:
#week1:
shapiro.test(aaaayang_week1$total_scfa) 
ggplot(aaaayang_week1, aes(x = total_scfa)) + geom_histogram() # p = 1.236e-05. Non-normal
qqnorm(aaaayang_week1$total_scfa, main='Inactive')
qqline(aaaayang_week1$total_scfa)

#week3:
shapiro.test(aaaayang_week3$total_scfa) 
ggplot(aaaayang_week3, aes(x = total_scfa)) + geom_histogram() # p = 0.2427. Normal
qqnorm(aaaayang_week3$total_scfa, main='Inactive')
qqline(aaaayang_week3$total_scfa)

#Variance:
aaaayang1 <- aaaayang %>%
  filter(supplement_consumed == "Inulin") %>%
  mutate(total_scfa = acetate_mmol_kg + butyrate_mmol_kg + propionate_mmol_kg) %>%
  na.omit(aaaayang1)
aaaayang1

bartlett.test(total_scfa ~ study_week, data = aaaayang1) # p = 0.05898. Below 0.1, so variances not really equal. 

#Evaluation: Normality and variance assumptions not satisfied, so Wilcoxon to be used. 

# test
wilcox.test(total_scfa ~ study_week, data = aaaayang1, paired = FALSE, alternative = "two.sided")

# result?
#p-value = 0.05793. This value is below 0.1, so I can say, reasonably, that there is a change in SCFA concentration from Week 1 to Week 3. 
```

Based on the results of the test, conclude if there was a difference in SCFAs between weeks. 

P-value = 0.05793. This value is below 0.1, so I can say, reasonably, that there is a change in SCFA concentration from Week 1 to Week 3. 

### Question 10

Is there a correlation between the abundace of hydrogen generators and breath hydrogen? Create a plot that shows the relationship and conduct the appropriate statistical test(s). [2 points]
```{r}
all_data_hydrogen <- all_data %>%
  filter(category == "hydrogen_gen")
all_data_hydrogen

# plot
hydrogen_plot <- all_data_hydrogen %>%
  ggplot(aes(x = relative_abundance,
              y = h2_mean)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +
  xlab("Mean Hydrogen Concentration") + 
  ylab("Relative Abundance of Hydrogen Generators") 
  ggsave(path = "Figures", filename = "hydrogen_plot.png")


# test
h2_test <- all_data_hydrogen %>%
  lm(h2_mean ~ relative_abundance, data = .)
summary(h2_test) #view results 

# result? p = <2e-16.  Multiple R-squared = 0.0001352. Adjusted R-squared = -0.0002808.
```

Based upon the low R-squared values and the low p-value, there appears to be no correlation between the Hydrogen concentration and the Relative Abundance of the Hydrogen Generators. There is however, clustering of the points towards lower mean hydrogen concentration values. 

Is there a correlation between the abundace of hydrogen consumers and breath methane? Create a plot that shows the relationship and conduct the appropriate statistical test(s). [2 points]
```{r}
# plot
all_data_cons <- all_data %>%
  filter(category == "hydrogen_consumers") 

all_data_cons %>%
  ggplot(aes(x = relative_abundance,
              y = ch4_mean)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +
  xlab("Mean Methane Concentration") + 
  ylab("Relative Abundance of Hydrogen Generators") 
  ggsave(path = "Figures", filename = "methane_plot.png")
  
all_data_cons <- all_data_cons %>%
  filter(category == "hydrogen_consumers")

# test
ch4_test <- all_data_cons %>%
  lm(ch4_mean ~ relative_abundance, data = .)
summary(ch4_test) #view results 

# result? p = <2e-16. Multiple R-squared = 0.2407. Adjusted R-squared = 0.2403. 
```

What can you conclude from these results?

Based upon the low p- and R-squared values, it can be said that there is little to no correlation between our two data groups of interest, although the relationship is still stronger than in the previous data sample (generators and hydrogen). 

### Extra credit:

Is the abundace of hydrogen producers and consumers related? Create a plot, check assumptions, conduct the appropriate statistical test(s), and interpret the results. [1 point]
```{r}
# plot


# test(s)
```

<type answer here> 


-----
end